---
title: "Coursework"
author: "Vinuji Hewapathirana"
date: "2025-03-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
set.seed(123) # For reproducibility
```

## Introduction

This report presents a simulation and analysis of a customer queuing system for a bank experiencing high customer footfall. The bank currently operates with two service counters and is considering adding a third counter if the average waiting time exceeds 15 minutes. The simulation aims to model the current system and predict performance with an additional counter.

## Methodology

### Queuing Model Formulation

The bank's queuing system is modeled with the following parameters:

- Customer arrivals follow a Poisson distribution (average rate: 10 customers/hour)
- Service times follow an Exponential distribution (average service time: 5 minutes/customer)
- Two service counters, each with a single server
- First-come, first-served (FCFS) queue discipline
- Simulation duration: 8-hour workday (480 minutes)

This can be classified as an M/M/2 queuing system where:
- M: Markovian (Poisson) arrival process
- M: Markovian (Exponential) service time distribution
- 2: Number of servers

### Simulation Implementation

```{r}
# Function to generate inter-arrival times (Poisson arrivals)
generate_arrivals <- function(rate, total_time) {
  # Convert rate from customers per hour to customers per minute
  rate_per_minute <- rate / 60
  
  # Generate inter-arrival times using exponential distribution
  inter_arrival_times <- rexp(1000, rate_per_minute)
  
  # Convert to actual arrival times
  arrival_times <- cumsum(inter_arrival_times)
  
  # Keep only arrivals within the simulation time
  arrival_times <- arrival_times[arrival_times <= total_time]
  
  return(arrival_times)
}

# Function to generate service times (Exponential distribution)
generate_service_times <- function(mean_service_time, num_customers) {
  # Generate service times
  service_times <- rexp(num_customers, 1/mean_service_time)
  
  return(service_times)
}

# Function to simulate the queuing system
simulate_queue <- function(arrival_times, service_times, num_servers) {
  num_customers <- length(arrival_times)
  
  # Initialize tracking variables
  waiting_times <- numeric(num_customers)
  service_start_times <- numeric(num_customers)
  service_end_times <- numeric(num_customers)
  
  # Track each server's availability time
  server_available_times <- rep(0, num_servers)
  
  # Track which server serves which customer
  server_assignment <- numeric(num_customers)
  
  # For each customer
  for (i in 1:num_customers) {
    # Find the earliest available server
    earliest_server <- which.min(server_available_times)
    
    # Record which server is serving this customer
    server_assignment[i] <- earliest_server
    
    # Calculate when service can start
    service_start_times[i] <- max(arrival_times[i], server_available_times[earliest_server])
    
    # Calculate waiting time
    waiting_times[i] <- service_start_times[i] - arrival_times[i]
    
    # Calculate service end time
    service_end_times[i] <- service_start_times[i] + service_times[i]
    
    # Update server availability
    server_available_times[earliest_server] <- service_end_times[i]
  }
  
  # Calculate system exit times
  exit_times <- service_end_times
  
  # Calculate queue lengths at different points in time
  time_points <- sort(unique(c(arrival_times, service_start_times)))
  queue_lengths <- numeric(length(time_points))
  
  for (t in 1:length(time_points)) {
    current_time <- time_points[t]
    in_queue <- sum(arrival_times <= current_time & service_start_times > current_time)
    queue_lengths[t] <- in_queue
  }
  
  # Calculate server utilization
  total_busy_time <- numeric(num_servers)
  for (s in 1:num_servers) {
    # Find all customers served by this server
    server_indices <- which(server_assignment == s)
    if (length(server_indices) > 0) {
      total_busy_time[s] <- sum(service_end_times[server_indices] - service_start_times[server_indices])
    } else {
      total_busy_time[s] <- 0
    }
  }
  
  # Calculate server utilization using max(exit_times) as total simulation time
  total_simulation_time <- max(exit_times)
  server_utilization <- sum(total_busy_time) / (num_servers * total_simulation_time)
  
  # Prepare results
  results <- list(
    waiting_times = waiting_times,
    queue_lengths = queue_lengths,
    time_points = time_points,
    server_utilization = server_utilization,
    average_waiting_time = mean(waiting_times),
    average_queue_length = mean(queue_lengths),
    max_queue_length = max(queue_lengths),
    total_customers = num_customers
  )
  
  return(results)
}

# Run the simulation
run_simulation <- function(arrival_rate, service_time, simulation_time, num_servers) {
  # Generate arrivals
  arrival_times <- generate_arrivals(arrival_rate, simulation_time)
  
  # Generate service times
  service_times <- generate_service_times(service_time, length(arrival_times))
  
  # Run simulation
  results <- simulate_queue(arrival_times, service_times, num_servers)
  
  return(results)
}
```

### Simulation Execution

```{r}
# Set parameters
arrival_rate <- 10  # customers per hour
service_time <- 5   # minutes per customer
simulation_time <- 480  # minutes (8-hour workday)

# Run simulation with 2 servers
results_2_servers <- run_simulation(arrival_rate, service_time, simulation_time, 2)

# Run simulation with 3 servers
results_3_servers <- run_simulation(arrival_rate, service_time, simulation_time, 3)
```

## Results and Analysis

### Current System Performance (2 Servers)

```{r}
# Display key metrics for 2 servers
cat("Performance with 2 servers:\n")
cat("Average waiting time:", round(results_2_servers$average_waiting_time, 2), "minutes\n")
cat("Average queue length:", round(results_2_servers$average_queue_length, 2), "customers\n")
cat("Maximum queue length:", results_2_servers$max_queue_length, "customers\n")
cat("Server utilization:", round(results_2_servers$server_utilization * 100, 2), "%\n")
cat("Total customers served:", results_2_servers$total_customers, "\n")
```

### Performance with Additional Server (3 Servers)

```{r}
# Display key metrics for 3 servers
cat("Performance with 3 servers:\n")
cat("Average waiting time:", round(results_3_servers$average_waiting_time, 2), "minutes\n")
cat("Average queue length:", round(results_3_servers$average_queue_length, 2), "customers\n")
cat("Maximum queue length:", results_3_servers$max_queue_length, "customers\n")
cat("Server utilization:", round(results_3_servers$server_utilization * 100, 2), "%\n")
cat("Total customers served:", results_3_servers$total_customers, "\n")
```

### Visualizations

```{r, fig.width=10, fig.height=5}
# Create dataframe for waiting time distribution
waiting_times_df <- data.frame(
  Waiting_Time = c(results_2_servers$waiting_times, results_3_servers$waiting_times),
  Servers = factor(c(rep("2 Servers", length(results_2_servers$waiting_times)), 
                   rep("3 Servers", length(results_3_servers$waiting_times))))
)

# Histogram of waiting times
ggplot(waiting_times_df, aes(x = Waiting_Time, fill = Servers)) +
  geom_histogram(position = "dodge", binwidth = 2, alpha = 0.7) +
  labs(title = "Distribution of Customer Waiting Times",
       x = "Waiting Time (minutes)",
       y = "Number of Customers") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

```{r, fig.width=10, fig.height=5}
# Create dataframe for queue lengths
queue_lengths_2 <- data.frame(
  Time = results_2_servers$time_points,
  Queue_Length = results_2_servers$queue_lengths,
  Servers = "2 Servers"
)

queue_lengths_3 <- data.frame(
  Time = results_3_servers$time_points,
  Queue_Length = results_3_servers$queue_lengths,
  Servers = "3 Servers"
)

queue_lengths_df <- rbind(queue_lengths_2, queue_lengths_3)

# Line plot of queue lengths
ggplot(queue_lengths_df, aes(x = Time, y = Queue_Length, color = Servers)) +
  geom_line() +
  labs(title = "Queue Length Over Time",
       x = "Time (minutes)",
       y = "Queue Length (number of customers)") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

### Performance Comparison

```{r}
# Create dataframe for comparison
comparison_df <- data.frame(
  Metric = c("Average Waiting Time (min)", "Average Queue Length", "Server Utilization (%)"),
  Two_Servers = c(round(results_2_servers$average_waiting_time, 2),
                 round(results_2_servers$average_queue_length, 2),
                 round(results_2_servers$server_utilization * 100, 2)),
  Three_Servers = c(round(results_3_servers$average_waiting_time, 2),
                   round(results_3_servers$average_queue_length, 2),
                   round(results_3_servers$server_utilization * 100, 2)),
  Improvement = c(round((results_2_servers$average_waiting_time - results_3_servers$average_waiting_time) / 
                      results_2_servers$average_waiting_time * 100, 2),
                round((results_2_servers$average_queue_length - results_3_servers$average_queue_length) / 
                      results_2_servers$average_queue_length * 100, 2),
                round((results_2_servers$server_utilization - results_3_servers$server_utilization) / 
                      results_2_servers$server_utilization * 100, 2))
)

knitr::kable(comparison_df, caption = "Performance Comparison Between 2 and 3 Servers")
```

## Discussion and Interpretation

The simulation results provide valuable insights into the bank's current queuing system and the potential impact of adding a third service counter.

With two service counters, our simulation shows that customers wait an average of approximately [insert value] minutes before being served. The average queue length is around [insert value] customers, with a maximum of [insert value] customers observed in the queue at peak times. The server utilization rate of [insert value]% indicates that the current servers are working at high capacity.

According to the bank's criteria, if the average waiting time exceeds 15 minutes, an additional counter should be considered. Our simulation shows that [does/does not] exceed this threshold.

When simulating the system with three service counters, the average waiting time decreases to approximately [insert value] minutes, representing a [insert value]% improvement. The average queue length also decreases to [insert value] customers, and server utilization drops to [insert value]%. This reduction in utilization suggests that servers would have more downtime, but it also provides flexibility to handle unexpected surges in customer arrivals.

The visualization of waiting times shows [describe pattern from histogram]. The queue length over time visualization illustrates [describe pattern from line chart]. These patterns indicate [interpretation of patterns].

## Conclusion and Recommendations

Based on the simulation results and analysis, I recommend that the bank [add/not add] a third service counter. This recommendation is based on the following observations:

1. The current average waiting time of [insert value] minutes [is/is not] above the bank's threshold of 15 minutes.
2. Adding a third counter would [significantly/marginally] improve customer experience by reducing waiting times by [insert value]%.
3. The trade-off between improved customer service and the cost of an additional counter should be carefully considered, as server utilization would decrease to [insert value]%.

Alternative strategies that the bank could consider include:
- Implementing an express lane for customers with simple transactions
- Adopting a digital queue management system to improve customer perception of waiting time
- Scheduling more staff during peak hours instead of adding a permanent third counter

In conclusion, the simulation provides evidence that [summarize main finding and recommendation]. Further simulations could explore these alternative strategies to optimize the bank's operations.